# Creating Longitudinal Datasets From Individual Records

```{r packages, include=F, echo=F, warning=F}
require("tidyverse", quietly=T, warn.conflicts = F)
require("Hmisc", quietly=T, warn.conflicts = F)
require("devtools", quietly=T, warn.conflicts = F)
require("Scotty", quietly=T, warn.conflicts = F)
mu <- markupSpecs$html  
require("knitr", quietly=T, warn.conflicts = F)
require("lubridate", quietly=T, warn.conflicts = F)
require("gridExtra", quietly=T, warn.conflicts = F)
```

## Introduction  

### Packages to use  

`r mu$session(cite=F)`  

I often come across the following issue in my work:  

  Sometimes you are working with a dataset where each row is a nursing home assessment, admission record or some other per person observation. However, perhaps you are more interesting in analyzing group-level changes over time. In order to do this, you need to reshape and summarize these individual records into counts in a "panel" dataset. In this new dataset structure I want each row to be a unique time- group- summary of the data. Some extensions of this include computing incidence (no. events per 1000 persons) and incidence density (no. events per 1000 person-years) measures. I will go through some examples and show how these datasets and measures can be constructed from person or observation unit-level data, assuming you had cohort entry-dates (i.e. admission), event dates (the date of some thing you wish to quantify) and stop-dates (i.e. discharge).  

## Construct dataset  

First I will construct a test dataset to use in this chapter and subsequent ones.  

In this hypothetical I take a group of admissions that starts counting on 2000/01/01. Each entry will have a random exit up to 1000 days from entry, but censored at 2009/12/31 (because in my hypothetical example this is my study endpoint). Each entry will then have a random group classification (state), and event (0 or 1).  

The event will be assumed one per admission (E.g. death..). I didn't allow for multiple admissions by person. I then generate the event as drawn from a random bernoulli distribution with probability $p_g$ Where $g$ is a group-specific effect randomly generated from a uniform distribution between 0 and 0.3.       

```{r TestData}
set.seed(12345) #So you get same result
sampsize = 20000
df <- data.frame(id=1:sampsize,
                 CohortEntry = sample(seq(as.Date('2000/01/01'), as.Date('2009/12/31'), by='day'), replace=T, sampsize))
#CohortIn, CohortOut, Group
df <- df %>%
    mutate(CohortExit = CohortEntry+sample(0:365,sampsize, replace=T)) %>% #CohortExit Date (up to 1000 days from start)
    mutate(State = sample(state.name, sampsize, replace=T)) #Random state for each group
df$CohortExit <- as.Date(sapply(df$CohortExit, function(x) min(x,as.Date('2009/12/31'))), origin=origin) #Censor at 'study end'

#Group Effect
  State <- as.data.frame(state.name)
  State$Effect <- runif(50, min=0, max=0.3) #Random effect by group

#Generate random event by group effect
  getReffect <- function(df, group) {
    p <- df[df$'state.name'==group,"Effect"]
    event = as.integer(rbernoulli(1, p = p))
    return(event)
  }
  
df$Event <- sapply(df$id, function(x) getReffect(State, df$State[x])) #Generate random event


#For more complicated procedures below, assign random event date between cohort start with event=0 to NA
  randomDate <- function(TimeIn, TimeOut, Event) {
    RDate <- sample(TimeIn:TimeOut, 1, replace=T)
    RDate <- ifelse(Event==0,NA,RDate)
    return(RDate)
  }
  
  df$EventDate = sapply(df$id, function(x) randomDate(df$CohortEntry[x], df$CohortExit[x], df$Event[x]))
  df$EventDate = as.Date(df$EventDate, origin=origin)
kable(head(df, n=10), align=c('c'))
```

## Show Data  

```{r EntryDate, fig.width=8, warnings=F}
#Scatter and Fitted Line 
p1 <- ggplot(data=df, aes(x=CohortEntry)) + 
  geom_histogram(aes(y = ..density..), binwidth = 4, fill=I("blue"), alpha=I(0.4)) +
  geom_density(col=2) +
  xlab("Cohort Entry") +
  theme_bw()

p2 <- ggplot(data=df, aes(x=CohortExit)) + 
  geom_histogram(aes(y = ..density..), binwidth = 4, fill=I("blue"), alpha=I(0.4)) +
  geom_density(col=2) +
  xlab("Cohort Exit") +
  theme_bw()
grid.arrange(p1, p2, nrow=1)
```
  
  A simple, even distribution to work with (cohort exit is even except for censored at study end date).  

## Simple group counts  

If you simply seek to count the total no. of records by groups this is simple.  

```{r ByGroup}
  dfState <- df %>%
    group_by(State) %>% #Tells dplyr to create grouped object, and then execute following at that unit
      summarise(Records = n()) #count individuals
  
  cat("Counts of Records by State")
  kable(head(dfState, n=10), align=c('c'))
```

Also, if you wish to see the quantity of some event, this is easy also:  

### Simple Event Counts  

```{r ByGroupEv}
  dfState <- df %>%
    group_by(State) %>% #Tells dplyr to create grouped object, and then execute following at that unit
      summarise(Events = sum(Event)) #count no of events
  
  cat("Counts of Events by State")
  kable(head(dfState, n=10), align=c('c'))
```

Note how the number of events is quite different by group, because we specified this above. If you wish to count records by time, this is still pretty easy, but you have to be more specific. For example, if I want to count the number of cohort entries by year, this is how:

## Cohort entries by year counts  

```{r ByYear}
#First make year var
df <- df %>%
  mutate(EntryYear = year(CohortEntry)) #year function from lubridate
  
#Second group by this var
dfGroup <- df %>% 
  group_by(EntryYear) %>%  
  summarise(Records = n(), Events = sum(Event)) #count individuals
  cat("Counts of Records by Cohort Entry Year")
  kable(head(dfGroup, n=10), align=c('c'))
```

Here you can see the number who enter the cohort by year, and among those how many events are observed.  

## Incidence

The next step will get a little trickier. Let's say we aren't interested in how many individuals entered/exited the cohort in a given year as above. Rather we want to identify how many patients are in the cohort during a specified period of time (e.g. year). This is the "population at risk". We wish to estimate the the number of events / population at risk. This is  [incidence](https://en.wikipedia.org/wiki/Incidence_(epidemiology)). So we need to take the "CohortEntry", "CohortExit" date variables and compute how many individuals were in the cohort in year 1, year 2 etc. What makes this tricky is that individuals don't start and stop at the same time and cross multiple time units (years in this case). 

Here is one method where I compute the no. of persons in the cohort in a given year, "incident" events and event rate:  

```{r cohortinc}
  #Create New Dataframe by Time Unit
  TimeMin <- min(year(df$CohortEntry)) #lowest time unit observed
  TimeMax <- max(year(df$CohortExit)) #highest time unit observed
  
  #This following sequence step is good, in case a certain year was skipped 
  #(i.e. no admits that year)
  dfTime <- TimeMin:TimeMax %>% #Sequence years
    as_tibble() %>%
    mutate(x2 = NA, x3 = NA)
    names(dfTime) <- c("Year", "Residents", "Inc. Events")
    
  
  #Write a time-interval program for Residence (assuming x is year)
  InCohort <-  function(x, TimeIn, TimeOut) {
    #Note that the following line works because of R vectorization
    count <- if_else(x>=year(TimeIn) & x<=year(TimeOut),1,0) #Test if x is TimeIn<=x<=TimeOut
    InCohortN <- sum(count) #Add up total people
    return(InCohortN) #return
  }
  
  #Write a time-interval program for Event
  IncEvent <-  function(x, Event, EventDate) {
    #Note that the following line works because of R vectorization
    events <- if_else(Event==1 & x==year(EventDate),1,0) #Added condition of event==1
    InCohortEvents <- sum(events) #Add up total events in that year
    return(InCohortEvents) #return
  }
  dfTime$Residents <- sapply(dfTime$Year, function(x) InCohort(x, df$CohortEntry,df$CohortExit))
  dfTime$'Inc. Events' <- sapply(dfTime$Year, function(x) IncEvent(x, df$Event, df$EventDate))
  dfTime$'Event Rate' <- dfTime$'Inc. Events' / dfTime$Residents
  kable(head(dfTime, n=10), align=c('c'), digits=3)
```

I didn't specify a time-trend in my sample generation, and that is why the event rate is relatively constant over time.  

We can double-check this worked with the following specific code:  

```{r checkInd}
  #Logic
  # IF 2004 is less than or equal to EntryDate (i.e. they entered before or during 2004)
  # AND 2004 is less than or equal to Exit (i.e. they exited after or during 2004)
  Check <- ifelse(2004>=year(df$CohortEntry) & 2004<=year(df$CohortExit),1,0)
  cat("2004 people :", sum(Check))
```

## Prevalence  

  The primary difference between incidence and prevalence is that incidence is only counting *new cases*. In order to compute prevalence, we need to identify individuals still in the cohort in a given year, but having the event in previous years as well.  
  
Here's how:  

```{r cohortPrev}
  #Code same as above
  TimeMin <- min(year(df$CohortEntry)) 
  TimeMax <- max(year(df$CohortExit)) 
  dfTime <- TimeMin:TimeMax %>% 
    as_tibble() %>%
    mutate(x2 = NA, x3 = NA)
    names(dfTime) <- c("Year", "Residents", "Prev. Events")
    
  InCohort <-  function(x, TimeIn, TimeOut) {
    count <- if_else(x>=year(TimeIn) & x<=year(TimeOut),1,0) 
    InCohortN <- sum(count) 
    return(InCohortN) 
  }
  
  #Write a time-interval program for Event
  PrevEvent <-  function(x, TimeIn, TimeOut, Event, EventDate) {
    #Key difference follows: 
    events <- if_else(x>=year(TimeIn) & x<=year(TimeOut) & Event==1 & x>=year(EventDate),1,0)
    InCohortEvents <- sum(events) #Add up total events in that year
    return(InCohortEvents) #return
  }
  
  dfTime$Residents <- sapply(dfTime$Year, function(x) InCohort(x, df$CohortEntry,df$CohortExit))
  dfTime$'Prev. Events' <- sapply(dfTime$Year, function(x) PrevEvent(x, df$CohortEntry, df$CohortExit, df$Event, df$EventDate))
  dfTime$'Prevalence Rate' <- dfTime$'Prev. Events' / dfTime$Residents
  kable(head(dfTime, n=10), align=c('c'), digits=3)
```

Note how the prevalence rate is higher, because you are counting events which happened in previous years. Also, the prevalence rate in the first year 2000 is the same as the incidence rate because we have no information before 2000 to add into the prevalent rate. If this is a problem in your empirical research, one approach would be to only count events from 2001 onward, and use the first year as a "lead-in" period.  

## Incidence Density  

The next goal is to report not the number of events per total population in a given time interval, but rather the number of events per person time. This is sometimes called [incidence density](https://en.wikipedia.org/wiki/Incidence_(epidemiology)), often reported as "x events per 100-person years". Conceptually you are saying you would on average expect x events in 1 person followed for 100 years, or x events in 100 persons followed for one year. This can be very useful when there is differing time accrued by the unit of observation (person, admission etc.). However, the relevance and interpretability of this measure is very specific to the thing being studied, i.e. it may not be a reasonable assumption that the event rate is constant over 100 years!  

The key new measure here is person-time, heres how to compute it:

```{r perTime}
df <- df %>%
    mutate(PerTime = as.integer(CohortExit - CohortEntry)) #timeDiff
#Second group by this var
dfGroup <- df %>% 
  group_by(EntryYear) %>%  
  summarise(Records = n(), 'Days in Cohort' = sum(PerTime)) #count individuals
  kable(head(dfGroup, n=10), align=c('c'))
```

That was easy! person- or unit-time accrued is simply `CohortExit - CohortEntry`. However, note we are reporting person-time by the year individuals entered the cohort, not an actual interval of time. It gets more complicated if you want to report aggregated person-time by some time interval. Because our individuals cross years, some of their person time accrues to one or more years.  

Here is how to do this:  

```{r perTimeByYear}
#Code same as above
  TimeMin <- min(year(df$CohortEntry)) 
  TimeMax <- max(year(df$CohortExit)) 
  dfTime <- TimeMin:TimeMax %>% 
    as_tibble() %>%
    mutate(x2 = NA, x3 = NA)
    names(dfTime) <- c('Year', 'Residents','Inc. Events')
    
  InCohort <-  function(x, TimeIn, TimeOut) {
    count <- if_else(x>=year(TimeIn) & x<=year(TimeOut),1,0) 
    InCohortN <- sum(count) 
    return(InCohortN) 
  }

  dfTime$Residents <- sapply(dfTime$Year, function(x) 
    InCohort(x, df$CohortEntry,df$CohortExit)) #Number of residents
  
  IncEvent <-  function(x, Event, EventDate) {
    #Note that the following line works because of R vectorization
    events <- if_else(Event==1 & x==year(EventDate),1,0) #Added condition of event==1
    InCohortEvents <- sum(events) #Add up total events in that year
    return(InCohortEvents) #return
  }
  
  dfTime$'Inc. Events' <- sapply(dfTime$Year, function(x) 
    IncEvent(x, df$Event, df$EventDate))

  #Function to compute person time
  TimeInCohort <- function(x, TimeIn, TimeOut) {
    
    #Key steps here:
    FirstDay <- as.Date(paste0('01/01/',x), format='%m/%d/%Y', origin=origin)
    LastDay <- as.Date(paste0('12/31/',x), format='%m/%d/%Y', origin=origin)
    
    #Compute starting point as either first day of year or TimeIn 
    #if TimeIn is > FirstDay
    #'' opposite for Year Stop
    YearStart <- sapply(TimeIn, function(x) max(x, FirstDay))
    YearStop <- sapply(TimeOut, function(x) min(x, LastDay))
    
    #Compute DaysInYear, if present in that year
    DaysInYear  <- ifelse(x>=year(TimeIn) & x<=year(TimeOut), YearStop - YearStart, 0)
    InCohortDays <- sum(DaysInYear) 
    return(InCohortDays)
  }  
  dfTime$'Person Time' <- sapply(dfTime$Year, function(x) 
    TimeInCohort(x, df$CohortEntry,df$CohortExit)) #Person time per year
  

  kable(head(dfTime, n=10), align=c('c'))
```

Then from this computing the incidence density rate is trivial, and you can report as some arbitrary quantity (i.e. per 100 years).  

```{r IncDensity}
#Code same as above
  #Events / Patient Days
  dfTime$`Incidence Density` <- dfTime$`Inc. Events` / dfTime$`Person Time`
  
  #Events / Patients Days * 365 days * 100 years
  dfTime$`Event rate per 100 person years` <- dfTime$`Incidence Density` * 365 * 100
  
  kable(head(dfTime, n=10), align=c('c'), digits = 4)
```

The intepretation "In 2003, the event rate 29.7 per 100 person-years" or "In 2003, the 1-year risk of an event is 29.7 per 100 persons".  