# Creating Longitudinal DataSets From Individual Records

```{r packages, include=F, echo=F}
require("tidyverse", quietly=T, warn.conflicts = F)
require("Hmisc", quietly=T, warn.conflicts = F)
require("devtools", quietly=T, warn.conflicts = F)
require("Scotty", quietly=T, warn.conflicts = F)
mu <- markupSpecs$html  
require("knitr", quietly=T, warn.conflicts = F)
```

### Packages to use  
`r mu$session(cite=F)`  

## Introduction  

I often come across the following issue in my work:  

  I have a dataset where each row is a nursing home assessment, admission record or some other per person observation. However, I need to reshape and summarize these individual records into counts in a panel dataset. In this new dataset structure I want each row to be a unique time- group- cross-section of the data. Some extensions of this include computing incidence (no. events per 1000 persons) and incidence density (no. events per 1000 person-years) measures. I will go through some examples and show how these datasets and measures can be constructed from person or observation unit-level data, assuming you had cohort entry-dates (i.e. admission), event dates (the date of some thing you wish to quantify) and stop-dates (i.e. discharge). 

## Construct dataset

```{r TestData}

```

## Show Data
