# Creating Longitudinal Datasets From Individual Records

```{r packages, include=F, echo=F, warning=F}
require("tidyverse", quietly=T, warn.conflicts = F)
require("Hmisc", quietly=T, warn.conflicts = F)
require("devtools", quietly=T, warn.conflicts = F)
require("Scotty", quietly=T, warn.conflicts = F)
mu <- markupSpecs$html  
require("knitr", quietly=T, warn.conflicts = F)
require("lubridate", quietly=T, warn.conflicts = F)
require("grid", quietly=T, warn.conflicts = F)
```

### Packages to use  
`r mu$session(cite=F)`  

## Introduction  

I often come across the following issue in my work:  

  Sometimes you are working with a dataset where each row is a nursing home assessment, admission record or some other per person observation. However, perhaps you are more interesting in analyzing group-level changes over time. In order to do this, you need to reshape and summarize these individual records into counts in a "panel" dataset. In this new dataset structure I want each row to be a unique time- group- summary of the data. Some extensions of this include computing incidence (no. events per 1000 persons) and incidence density (no. events per 1000 person-years) measures. I will go through some examples and show how these datasets and measures can be constructed from person or observation unit-level data, assuming you had cohort entry-dates (i.e. admission), event dates (the date of some thing you wish to quantify) and stop-dates (i.e. discharge).  

## Construct dataset  

First I will construct a dataset of admissions that starts counting on 2000/01/01. Each entry will have a random exit up to 1000 days from entry, but censored at 2009/12/31 (because in my hypothetical example this is my study endpoint). Each entry will then have a random group classification (state), and event (0 or 1).    
```{r TestData}
df <- data.frame(id=1:10000,
                 CohortEntry = sample(seq(as.Date('2000/01/01'), as.Date('2009/12/31'), by='day'), replace=T, 10000))
df <- df %>%
    mutate(CohortExit = CohortEntry+sample(0:365,10000, replace=T)) %>% #CohortExit Date (up to 1000 days from start)
    mutate(Group = sample(state.name, 10000, replace=T)) %>% #Random state for each group
    mutate(Event = sample(0:1, 10000, replace=T))

df$CohortExit <- as.Date(sapply(df$CohortExit, function(x) min(x,as.Date('2009/12/31'))), origin=origin) #Censor at 'study end'
```

## Show Data
```{r EntryDate, warnings=F}
#Scatter and Fitted Line 
p1 <- ggplot(data=df, aes(x=CohortEntry)) + 
  geom_histogram(aes(y = ..density..), binwidth = 5, fill=I("blue"), alpha=I(0.4)) +
  geom_density(col=2) +
  xlab("Cohort Entry") +
  theme_bw()

p2 <- ggplot(data=df, aes(x=CohortExit)) + 
  geom_histogram(aes(y = ..density..), binwidth = 5, fill=I("blue"), alpha=I(0.4)) +
  geom_density(col=2) +
  xlab("Cohort Exit") +
  theme_bw()
grid.arrange(p1, p2, nrow=1)
```
  A simple, even distribution to work with (cohort exit is even except censored at study end date).  
  
## Reshape Process

If you simply seek to count the total no. of records by groups this is simple. 

### Simple group counts
```{r ByGroup}
  dfGroup <- df %>%
    group_by(State) %>% #Tells dplyr to create grouped object, and then execute following at that unit
      summarize(Records = n()) #count individuals
  cat("Counts of Records by State")
  head(dfGroup, n=10)
```
If you wish to count records this is about as simple, but you have to be more specific. For example, if I want to count the number of Cohort entries by year, this is how:

### Cohort entries by year counts
```{r ByYear}
#First make year var
df <- df %>%
  mutate(EntryYear = year(CohortEntry)) #year function from lubridate
  
#Second group by this var
dfGroup <- df %>%    
  group_by(EntryYear) %>%       
  summarize(Records = n()) #count individuals  
  cat("Counts of Records by Cohort Entry Year")  
  head(dfGroup, n=10)
```



### Calculate time distance  
```{r newVars}
df <- df %>%
    mutate(TimeDiff = as.integer(CohortExit - CohortEntry)) #timeDiff
```
