# Heteroskedastic & Cluster Robust Standard Errors  

```{r packages, include=T, echo=F}
require("Scotty")
require("tidyverse")
require("Hmisc")
mu <- markupSpecs$html  

require("sandwich")
require("lmtest")
require("boot")
```

## Introduction 

In this chapter we are evaluating R's capability to compute standard errors. Overall like most things, R has substantial capacity but its flexibility can be daunting. To bring this flexibility down to earth, I lay out the background, provide practical recommendations, user-written commands and benchmark to STATA. 

### Packages to use
`r mu$session()`

"Scotty" is my own package. "tidyverse" is Wickam et al. general suite of packages/commands to work with R. "Hmisc" is Frank Harrel's miscellaneous commands, many of which are quite useful. 

"sandwich", "lmtest" and "boot" are specifically relevant to this chapter in order to compute various standard errors (SE). 

### Test Data

The cluster data was obtained from:
http://www.kellogg.northwestern.edu/faculty/petersen/htm/papers/se/test_data.txt
http://www.kellogg.northwestern.edu/faculty/petersen/htm/papers/standarderror_extra_tables.pdf
https://sites.google.com/site/waynelinchang/r-code
http://www.kellogg.northwestern.edu/faculty/petersen/htm/papers/se/test_data.htm
https://thetarzan.wordpress.com/2011/05/28/heteroskedasticity-robust-and-clustered-standard-errors-in-r/

### Other References
## R's calculation of standard errors

## Heteroskedascity

## Clustering

## Stata comparison

A full discussion of STATA programming can be seen here: http://www.kellogg.northwestern.edu/faculty/petersen/htm/papers/se/se_programming.htm
STATA blog:http://www.stata.com/support/faqs/statistics/standard-errors-and-vce-cluster-option/

Briefly: In Stata one can specify a variance-covariance matrix that is heteroskedastic consistent with the *vce(robust)* option in regression models.  


e.g. STATA example
```{}
regress y x z, vce(robust)
```

A Huber-White variance-covariance matrix can also be computed by some group with the **vce(cluster *group*)** option in regression models.

e.g. STATA example
```{}
regress y x z, vce(cluster(group))
```

See: http://www.stata.com/support/faqs/statistics/standard-errors-and-vce-cluster-option/
## Heteroskedastic consistent errors in R  

## Cluster robust errors in R  

## Block bootstrapping  

An alternative to computing special variance-covariance matrices is non-parametric  "block" bootstrapping. To do this, you perform a bootstrapping procedure where you sample the group or "block" instead of unit observation. This has been shown to be about as consistent and unbiased as the above sandwich estimators, and may be advantgeous when the number of clusters is small.(Insert Bertrand citation).  

### Bootstrap Program
```{r BootProg}
Boot.ATE <- function (model, treat, R = 250, block = "", df) 
{
  require(boot)
  require(dplyr)
  family <- model$family
  if (block == "") {
    boot.mod <- function(x, i, model, treat) {
      samp.df <- x[i, ]
      samp.glm <- try(glm(model, data = samp.df, family = family))
      if (inherits(samp.glm, "try-error")) {
        coef <- NA
        ate <- NA
        rr <- NA
        c(coef, ate, rr)
      }
      else {
        df2 <- samp.df
        df2[, paste(treat)] = 1
        pred1. <- predict.glm(samp.glm, newdata = df2, 
          type = "response")
        df2[, paste(treat)] = 0
        pred0. <- predict.glm(samp.glm, newdata = df2, 
          type = "response")
        coef <- samp.glm$coefficients[paste0(treat)]
        ate <- mean(pred1.) - mean(pred0.)
        rr <- mean(pred1.)/mean(pred0.)
        c(coef, ate, rr)
      }
    }
    boot.m <- boot(data = df, statistic = boot.mod, R = R, 
      model = model, treat = treat)
  }
  else {
    Groups = unique(df[, paste(block)])
    boot.mod <- function(x, i, model, treat, df, block, 
      iter = 0) {
      block.df <- data.frame(group = x[i])
      names(block.df) = block
      samp.df <- left_join(block.df, df, by = block)
      samp.glm <- try(glm(model, data = samp.df, family = family))
      if (inherits(samp.glm, "try-error")) {
        coef <- NA
        ate <- NA
        rr <- NA
        c(coef, ate, rr)
      }
      else {
        df2 <- samp.df
        df2[, paste(treat)] = 1
        pred1. <- predict.glm(samp.glm, newdata = df2, 
          type = "response")
        df2[, paste(treat)] = 0
        pred0. <- predict.glm(samp.glm, newdata = df2, 
          type = "response")
        coef <- samp.glm$coefficients[paste0(treat)]
        ate <- mean(pred1.) - mean(pred0.)
        rr <- mean(pred1.)/mean(pred0.)
        c(coef, ate, rr)
      }
    }
    boot.m <- boot(data = Groups, statistic = boot.mod, 
      R = R, model = model, treat = treat, df = df, block = block)
  }
  m1.confint <- c(model$coefficients[paste0(treat)], confint(model, 
    treat, level = 0.95))
  coeff = boot.ci(boot.m, index = 1, type = "perc")
  coeff = c(median(boot.m$t[, 1]), coeff$percent[, 4], coeff$percent[, 
    5])
  names(coeff) <- c("Coeff.", "2.5%", "97.5%")
  ate = boot.ci(boot.m, index = 2, type = "perc")
  ate = c(median(boot.m$t[, 2]), ate$percent[, 4], ate$percent[, 
    5])
  names(ate) <- c("ATE", "2.5%", "97.5%")
  rr = boot.ci(boot.m, index = 3, type = "perc")
  rr = c(median(boot.m$t[, 3]), rr$percent[, 4], rr$percent[, 
    5])
  names(rr) <- c("Rr", "2.5%", "97.5%")
  boot.iter = boot.m$t
  res = list(level = 0.95, model_ci = m1.confint, coeff = coeff, 
    ate = ate, rr = rr, boots = boot.iter)
  return(res)
}
```

## Acknowledgements
 This chapter is heavily adapted from several StackExchange and other blog posts.
 See:
 http://www.richard-bluhm.com/clustered-ses-in-r-and-stata-2/
 

## Bibliography
  
